The Two Dualities of Computation: Negative and Fractional Types

### 2012-02-29 rational.tex
changed to fractions

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev
%! Gx = \Gamma^{\times}
%! G = \Gamma
%! [] = \Box
%! |-->* = \mapsto^{*}
%! |-->> = \mapsto_{\ggg}
%! |--> = \mapsto
%! |- = \vdash
%! ==> = \Longrightarrow
%! <== = \Longleftarrow
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! ~> = \leadsto
%! -o = \multimap
%! ::= = &::=&
%! /= = \neq
%! @@ = \mu
%! forall = \forall
%! exists = \exists
%! empty = \epsilon
%! langRev = \Pi
%! langRevT = \Pi^{o}
%! langRevEE = \Pi^{\eta\epsilon}
%! theseus = Theseus
%! * = \times

@@

%subcode{bnf} include main
% Value Types, b = ... | b^^^
% Values, v = ... | v^^^
%
% Isomorphisms, iso &=& ... | eta* | eps*

@@

%subcode{opsem} include main
% eta* &: 1 <-> b^^^ * b :& eps*

%subcode{proof} include main
%@ |- v : b
%@@ |- v^^^ : b^^^

As with additive duals, we represent {{eta*}} and {{eps*}] using
`U'-shaped connectors. In the general case the wires for {{1}} will be
dropped as it can freely introduced or eliminated using {{uniti}} and
{{unite}}.

@@

The usual interpretation of {{b1*b2}} we have both a value of type
{{b1}} and a value of type {{b2}}. This interpretation is maintained
in the precense of fractionals. Hence an {{eta* : 1 <-> b^^^ * b}} is
to be viewed as a fission point for a value of type {{b}} and its
multiplicative inverse {{b^^^}}. This naturally raises the question --
which value of type {{b}} is produced? The isomorphism {{eta*}} does
not produce any specific value of type {{b}}, instead it produces a
place holder for a value such that the value of type {{b^^^}} is
always the dual of the chosen one.

@@

\paragraph*{Unification.}
A computational interpretation of this is that {{langRevEE}} acts like
a logic programming system and {{eta*}} acts as the site of creation
of a new typed logic variable of type {{b}} and its corresponding dual
of type {{b^^^}}. These two values are free to move independantly
through a circuit, but any operations on value affect what operations
are possible on the other since they are both related by an underlying
logic variable. Dually, {{eps*}} acts as a unification site for values
flowing along both branches. If the values are not exactly duals, then
the system fails to unify and hence the program is stuck. If the
values are exactly duals, then unification succeeds and {{eta*}}
returns {{1}}.

@@

In this context, let us go back to the thought that recursive such as
{{nat=nat+1}} have no meaningful solutions, whereas trees do. But the
interesting solution of trees is precisely that they involve
irrational numbers. Our usual notation for irrationals such as
{{2^{1/2} }} can actually be viewed as a thunk representing the
infinite computation {{1.4142..}}. Irrationals in this sense have a
very different information theoretic interpretation from full real
numbers. With Chaitin's Omega numbers he postulates the existence of
real numbers that have an infinite amount of information -- ones for
example are the solution to the question about the halting properties
of every Turning machine.

Real numbers in that sense really are a series of bits of unknown
information theoretic interpretation -- a sort of {{top}}. Numbers
such as {{2^{1/2} }} are very different in the sense that even though
they contain an infinite sequence of bits, there is a finite program
that generates those bits and its only the cost of computing bits that
is infinite.  The consistency of arithmetic isn't affected by
extending the rational field by adding any specific computable
irrational. So it might be possible to design a model of computing
where we can indeed deal with them. Such a model will have to explicit
thunks for the delayed computation represented by trees and never
equate and cancel two infinite computations unless the represent
exactly the same infinite computation.

>>>>

%subcode-inline{bnf-inline} regex \{\{(((\}[^\}])|[^\}])*)\}\} name main include langRev
%! Gx = \Gamma^{\times}
%! G = \Gamma
%! [] = \Box
%! |-->* = \mapsto^{*}
%! |-->> = \mapsto_{\ggg}
%! |--> = \mapsto
%! |- = \vdash
%! ==> = \Longrightarrow
%! <== = \Longleftarrow
%! <=> = \Longleftrightarrow
%! <-> = \leftrightarrow
%! ~> = \leadsto
%! -o = \multimap
%! ::= = &::=&
%! /= = \neq
%! @@ = \mu
%! forall = \forall
%! exists = \exists
%! empty = \epsilon
%! langRev = \Pi
%! langRevT = \Pi^{o}
%! langRevEE = \Pi^{\eta\epsilon}
%! theseus = Theseus
%! sqrt(x) = \sqrt{#x}
%! surd(p,x) = \sqrt[#p]{#x}
%! inv(x) = \frac{1}{#x}
%! frac(x,y) = \frac{#x}{#y}
%! * = \times

@@

%subcode{bnf} include main
% Value Types, b = ... | inv(b)
% Values, v = ... | inv(v)
%
% Isomorphisms, iso &=& ... | eta* | eps*

@@

%subcode{opsem} include main
% eta* &: 1 <-> inv(b) * b :& eps*

%subcode{proof} include main
%@ |- v : b
%@@ |- inv(v) : inv(b)

As with additive duals, we represent {{eta*}} and {{eps*}} using
`U'-shaped connectors. In the general case the wires for {{1}} will be
dropped as it can freely introduced or eliminated using {{uniti}} and
{{unite}}.

@@

The usual interpretation of {{b1*b2}} we have both a value of type
{{b1}} and a value of type {{b2}}. This interpretation is maintained
in the precense of fractionals. Hence an {{eta* : 1 <-> inv(b) * b}} is
to be viewed as a fission point for a value of type {{b}} and its
multiplicative inverse {{inv(b)}}. This naturally raises the question --
which value of type {{b}} is produced? The isomorphism {{eta*}} does
not produce any specific value of type {{b}}, instead it produces a
place holder for a value such that the value of type {{inv(b)}} is
always the dual of the chosen one.

@@

\paragraph*{Unification.}
A computational interpretation of this is that {{langRevEE}} acts like
a logic programming system and {{eta*}} acts as the site of creation
of a new typed logic variable of type {{b}} and its corresponding dual
of type {{inv(b)}}. These two values are free to move independantly
through a circuit, but any operations on value affect what operations
are possible on the other since they are both related by an underlying
logic variable. Dually, {{eps*}} acts as a unification site for values
flowing along both branches. If the values are not exactly duals, then
the system fails to unify and hence the program is stuck. If the
values are exactly duals, then unification succeeds and {{eta*}}
returns {{1}}.

@@

In this context, let us go back to the thought that recursive such as
{{nat=nat+1}} have no meaningful solutions, whereas trees do. But the
interesting solution of trees is precisely that they involve
irrational numbers. Our usual notation for irrationals such as
{{sqrt(2) }} can actually be viewed as a thunk representing the
infinite computation {{1.4142..}}. Irrationals in this sense have a
very different information theoretic interpretation from full real
numbers. With Chaitin's Omega numbers he postulates the existence of
real numbers that have an infinite amount of information -- ones for
example are the solution to the question about the halting properties
of every Turning machine.

Real numbers in that sense really are a series of bits of unknown
information theoretic interpretation -- a sort of {{top}}. Numbers
such as {{sqrt(2) }} are very different in the sense that even though
they contain an infinite sequence of bits, there is a finite program
that generates those bits and its only the cost of computing bits that
is infinite.  The consistency of arithmetic isn't affected by
extending the rational field by adding any specific computable
irrational. So it might be possible to design a model of computing
where we can indeed deal with them. Such a model will have to explicit
thunks for the delayed computation represented by trees and never
equate and cancel two infinite computations unless the represent
exactly the same infinite computation.
